Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-30 Manuscript/SystBiolSubmission/TEM_draft.tex
removed hypref

We inferred the morphological matrix using the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see \hyperref[SupplementaryMaterial]{online Appendix 1} and Fig. S1). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. We used the following priors for each tree: (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took \texttildelow 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the \hyperref[SupplementaryMaterial]{online Appendix 1}.

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{online Appendix 2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by dividing the Robinson-Foulds distance by the mean Robinson-Foulds distance from 1000 random tree comparisons \citep{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{online Appendix 2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{online Appendix 2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions (\citealt{Bhattacharyya}; see \hyperref[SupplementaryMaterial]{online Appendix 2} for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

\subsection{Individual effects of missing data parameters}
\subsubsection{Missing data for living taxa ($M_{L}$)}
When the number of living taxa with morphological data ($M_{L}$) decreases, entire rows of data are being removed from the living taxa part of the matrix. Because living taxa still have molecular characters available for phylogenetic inference (see \hyperref[Generating_the_matrix]{methods}),even if they have no morphological data, the relationships among them will always be fairly well-resolved (depending on the phylogenetic signal from the molecular part of the matrix). However, this missing data parameter has a huge influence on the placement of fossil taxa because a decrease in the $M_{L}$ parameter reduces the amount of overlapping data among the living and fossil taxa, meaning there is no part of the living taxa tree that the fossils can branch off.

@@

>>>>

We inferred the morphological matrix using the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see online Appendix 1 and Fig. S1). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. We used the following priors for each tree: (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took \texttildelow 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the online Appendix 1.

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see online Appendix 2 for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by dividing the Robinson-Foulds distance by the mean Robinson-Foulds distance from 1000 random tree comparisons \citep{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see online Appendix 2 for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see online Appendix 2 for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions (\citealt{Bhattacharyya}; see online Appendix  for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

\subsection{Individual effects of missing data parameters}
\subsubsection{Missing data for living taxa ($M_{L}$)}
When the number of living taxa with morphological data ($M_{L}$) decreases, entire rows of data are being removed from the living taxa part of the matrix. Because living taxa still have molecular characters available for phylogenetic inference (see mMthods),even if they have no morphological data, the relationships among them will always be fairly well-resolved (depending on the phylogenetic signal from the molecular part of the matrix). However, this missing data parameter has a huge influence on the placement of fossil taxa because a decrease in the $M_{L}$ parameter reduces the amount of overlapping data among the living and fossil taxa, meaning there is no part of the living taxa tree that the fossils can branch off.

@@

\end{document}
